---
title: "Reproducible Research: Peer Assessment 1"
author: "Allwyn Joseph"
date: "2/19/2017"
output: md_document
---
## Loading and preprocessing the data

Let's first load the data and perform some minor transformations. 
```{r simulation, echo=TRUE,message=FALSE}
library("lubridate")
library("dplyr")
fit_dat<- read.csv("activity.csv",header = TRUE, sep = "," )
fit_dat$date<- ymd(fit_dat$date)
```

## Mean total number of steps taken per day

Let's first create a dataframe 'summary' containing the total number of steps taken per day. 
```{r simulationdata, echo=TRUE,message=FALSE}
summary <- aggregate(fit_dat$steps, 
                   by=list(fit_dat$date), 
                   FUN=sum, 
                   na.rm = TRUE)
colnames(summary)<- c("Date","Total_Steps")
```

Below is a histogram displaying the total number of steps taken per day. 
```{r histogram, fig.height=5, echo=TRUE}
hist(rep(summary$Date,summary$Total_Steps), 
     breaks = 61, 
     freq  = TRUE, 
     xlab = "Days",
     ylab = "Total number of steps taken", 
     main = "Total number of steps taken per day")
```

Calculating the mean and median of the total number of steps taken:
```{r computetime, echo=TRUE,message=FALSE}
me<- mean(summary$Total_Steps)
md<- median(summary$Total_Steps)
```
The mean and median of the total number of step taken is `r me` and `r md` respectively. 
 
## Average daily activity pattern 

Let's create a dataframe 'time'and store the average steps on an interval basis.
```{r simulatedata, echo=TRUE,message=FALSE}
time<- data.frame(tapply(fit_dat$steps, 
                      fit_dat$interval, 
                      FUN = mean, 
                      na.rm = TRUE))

time$interval<- rownames(time) %>% as.numeric(time$interval)
colnames(time)<- c("Average","interval")
```

Below is the time series plot displaying the the average daily activity pattern:
```{r scatterplot, fig.height=5, echo=TRUE}
plot(time$interval,time$Average,
     type = "l", 
     xlab = "5 minute Intervals",
     ylab = "Number of Steps (Average)", 
     main = "Average Daily Activity Pattern")
```

Calculating the 5 minute interval with maximum number of steps:
```{r compute, echo=TRUE,message=FALSE}
int<- time[which.max(time$Average),2]
a<- int/5
```
The interval `r int` contains the maximum number of steps across all days, which equals to the `r a`th 5 minute interval.

## Inputing missing values

Calculating the total number of missing values:
```{r computetot, echo=TRUE,message=FALSE}
tot<- sum(is.na(fit_dat))
```
The total number of missing values is `r tot`

Computing the strategy for filling in the missing values:
```{r , echo=TRUE,message=FALSE}
miss <- aggregate(fit_dat$steps, 
                     by=list(fit_dat$date), 
                     FUN= mean, 
                     na.rm = TRUE)
colnames(miss)<- c("Date","Mean_Steps")
miss[is.na(miss)] <- 0
```

Forming a new dataframe without NA values:
```{r simulate, echo=TRUE,message=FALSE}
new_fit_dat<- fit_dat
new_fit_dat$steps <- ifelse(is.na(new_fit_dat$steps) == TRUE, 
                            time$Average[time$interval %in% new_fit_dat$interval], 
                            new_fit_dat$steps)
```

Plotting and comparing new and old data sets:
```{r plot, fig.height=5, fig.width=10, echo=TRUE}
par(mfrow =  c(1,2))

# old histogram
summary <- aggregate(fit_dat$steps, 
                     by=list(fit_dat$date), 
                     FUN=sum, 
                     na.rm = TRUE)
colnames(summary)<- c("Date","Total_Steps")
hist(rep(summary$Date,summary$Total_Steps), 
     breaks = 90, 
     freq  = TRUE, 
     xlab = "Days",
     ylab = "Steps taken", 
     main = "Total number of steps taken per day")

# new histogram
summary_new <- aggregate(new_fit_dat$steps, 
                     by=list(new_fit_dat$date), 
                     FUN=sum, 
                     na.rm = TRUE)
colnames(summary_new)<- c("Date","Total_Steps")
hist(rep(summary_new$Date,summary_new$Total_Steps), 
     breaks = 90, 
     freq  = TRUE, 
     xlab = "Days",
     ylab = "Steps taken", 
     main = "Total number of steps taken per day \nafter NA values were replaced")
n_me<- mean(summary_new$Total_Steps)
n_md<- median(summary_new$Total_Steps)

```
The new mean and new median of the total number of step taken is `r n_me` and `r n_md` respectively.
Both the mean and the median values have increased. Also, using the strategy of replace the missing values with the mean of 5 minute intervals have filled up the plot resulted in a credible and unexaggerated output.

## Patterns between weekdays and weekends

Creating a new factor variable in the dataset with two levels – 'weekday' and 'weekend' 
```{r datas, echo=TRUE, message=FALSE}
new_fit_dat$week <- ifelse(weekdays(new_fit_dat$date) %in% c("Saturday", "Sunday"), 
                           "weekend", 
                           "weekday")
miss[is.na(miss)] <- 0
```

Plotting and comparing weekday and weekend data sets:
```{r plots, fig.height=5, echo=TRUE}
library("ggplot2")
# grouping data on weekend, weekday and interval basis
group_dat<- group_by(new_fit_dat,interval,week)
group_dat<- summarise(group_dat, avg = mean(steps))

# plotting and comparing data
qplot(interval, avg, data=group_dat,
      geom="line",
      xlab="Interval",
      ylab="Number of Steps (Average)",
      main=" Comparison of Average steps on Weekends and Weekdays",
      facets =week ~ .)
```
From the graphs we understand that the level of activity is a little more evenly distributed in the case of weekends, where as during weekdays there is an initial spike followed by a relative drop in activity over the following intervals. 



